Optimal updating magnitude in adaptive flat-distribution-sampling simulations

### 2017-08-29 doc/invt.tex
add proof for alpha(t) > 1/t

As shown in Fig. \ref{fig:lj_alpha},
the optimal schedule for the Gaussian updating scheme
was more complex than the inverse-time one,
and the former tended to give a larger updating magnitude.
%
Since our equilibration run was sufficiently long,
the initial updating magnitude, $\alpha(0)$,
was around the ideal value of $a_0/2 = 5\times10^{-5}$
given by Eq. \eqref{eq:half_alpha0}.

@@

Here we give a geometric characterization
of the optimal schedule.
%
First, we define a normalized mass distribution as
%
\begin{equation}
  m(\bar q)
  =
  \frac{
    M(\bar q)
  }
  {
    \int_0^{ q(T) } M(q') \, d q'
  }
  =
  \frac{
    M(\bar q)
  }
  {
    C_M
  }
  ,
\notag
%\label{eq:mass_distr}
\end{equation}
%
such that
$\int_0^{q(T)} m(\bar q) \, d\bar q = 1$.
%
Then from Eqs. \eqref{eq:Lagrangian_const} and \eqref{eq:q_opt},
we find that for $\bar q(t) \equiv q(T) - q(t)$,
%
\begin{align}
  m\left( \bar q(t) \right)
  &=
  \frac{ 1 }
       { T \, \alpha(t) }
  ,
  \label{eq:mQ_invTa}
  \\
  \int_{\bar q(t)}^{ q(T) }
    m(q') \, d q'
  &=
  \frac t T
  .
  \label{eq:intmQ_tT}
\end{align}
%
This allows each point on the schedule,
$\bigl(t, \alpha(t)\bigr)$,
to be mapped to a point,
$\Bigl(\bar q(t), m\bigl(\bar q(t)\bigr)\Bigr)$,
on the mass distribution,
such that the ordinate of the latter curve
is $1/(T\alpha)$,
and the area under the curve in the domain, $[\bar q(t), q(T)]$,
is equal to $t/T$,
as shown in Fig. \ref{fig:massq}(a).

@@

>>>>

As shown in Fig. \ref{fig:lj_alpha},
the optimal schedule for the Gaussian updating scheme
was more complex than the inverse-time one,
and the former tended to give a larger updating magnitude,
in agreement with the discussion in Appendix \ref{sec:schedule_geometry}.
%
Since our equilibration run was sufficiently long,
the initial updating magnitude, $\alpha(0)$,
was around the ideal value of $a_0/2 = 5\times10^{-5}$
given by Eq. \eqref{eq:half_alpha0}.

@@

Here we give a geometric characterization
of the optimal schedule.
%
First, we define a normalized mass distribution as
%
\begin{equation}
  m(\bar q)
  =
  \frac{
    M(\bar q)
  }
  {
    \int_0^{ q(T) } M(q') \, d q'
  }
  =
  \frac{
    M(\bar q)
  }
  {
    C_M
  }
  ,
  %\notag
  \label{eq:mass_distr}
\end{equation}
%
such that
$\int_0^{q(T)} m(\bar q) \, d\bar q = 1$.
%
Then from Eqs. \eqref{eq:Lagrangian_const} and \eqref{eq:q_opt},
we find that for $\bar q(t) \equiv q(T) - q(t)$,
%
\begin{align}
  m\left( \bar q(t) \right)
  &=
  \frac{ 1 }
       { T \, \alpha(t) }
  ,
  \label{eq:mQ_invTa}
  \\
  \int_{\bar q(t)}^{ q(T) }
    m(\bar q) \, d \bar q
  &=
  \frac t T
  .
  \label{eq:intmQ_tT}
\end{align}
%
This allows each point on the schedule,
$\bigl(t, \alpha(t)\bigr)$,
to be mapped to a point,
$\Bigl(\bar q(t), m\bigl(\bar q(t)\bigr)\Bigr)$,
on the mass distribution,
such that the ordinate of the latter curve
is $1/(T\alpha)$,
and the area under the curve in the domain, $[\bar q(t), q(T)]$,
or the complementary cumulative distribution function of $m(\bar q)$,
is equal to $t/T$,
as shown in Fig. \ref{fig:massq}(a).

@@

We can use Eq. \eqref{eq:intmQ_tT} to show that
for two simulations with equal lengths and initial updating magnitudes,
the simulation under the Gaussian scheme with the optimal schedule
has a larger updating magnitude than
the value given by the inverse-time formula,
which is optimal for the single-bin scheme.
%
From Eqs. \eqref{eq:mass_func} and \eqref{eq:mass_distr},
we have
\begin{align}
  -\frac{ d \ln m(\bar q) }{ d \bar q }
  %=
  %-\frac{ dM^2(\bar q)/d\bar q } { 2 M^2(\bar q) }
  =
  \frac{ \sum_{k=1}^{n-1} \Gamma_k \, \lambda_k^3 \, e^{-2\lambda_k \bar q} }
  { \sum_{k=1}^{n-1} \Gamma_k \, \lambda_k^2 \, e^{-2\lambda_k \bar q} }
  \equiv
  \bar\lambda(\bar q)
  ,
  \label{eq:dlnmdq}
\end{align}
which is a weighted average of $\lambda_k$'s
with the weight being $\Gamma_k \, \lambda_k^2 \, e^{-2\lambda_k \bar q}$.
%
Let us use Eqs. \eqref{eq:intmQ_tT} and \eqref{eq:dlnmdq}
to compare the time of
reaching a certain updating magnitude, $\alpha_1$,
in the two simulations:
%
\begin{align}
  \frac{t}{T}
  %=
  %\int_{\bar q(t)}^{q(T)} m(\bar q) \, d\bar q
  =
  \int_{m( \bar q(t) )}^{m( q(T) )}
  \!\!\frac{ m(\bar q) } { dm(\bar q)/d\bar q } \, d m
  =
  \int^{m( \bar q(t) )}_{m( q(T) )}
  \!\!\frac{ d m } { \bar \lambda(\bar q) }
  ,
  \label{eq:t_int}
\end{align}
with the lower and upper limits,
$m\bigl(q(T)\bigr) = 1/\bigl[T\alpha(0)\bigr]$ and
$m\bigl(\bar q(t)\bigr) = 1/(T\alpha_1)$,
being the same in both cases.
%
Since for the Gaussian scheme,
we have $\bar \lambda(\bar q) \le 1$,
by Eq. \eqref{eq:lambda_Gaussian},
the integral in Eq. \eqref{eq:t_int}
is no less than that for the single-bin scheme,
i.e. it takes longer for the Gaussian scheme
than the single-bin scheme
to lower the updating magnitude
to the same level.
